New lower bounds for Schur and weak Schur numbers

### 2021-04-15 article/weakschur.tex
synthaxe
restructuration + indicatrice plus jolie

In this section, we generalize Rowley's constructions in \cite{RowleyWS}. We then introduce, by analogy with the third
section, the
integer \(\WS^+(n)\)
to build suitable templates.

@@

Up to now, there was no equivalent for weak Schur numbers of Abott and Hanson's construction \cite{AbbottHanson}. Here
we
give a general lower bound for weak Schur numbers as a function of both regular and weak Schur numbers.
The following theorem, inspired by Rowley's inequalities for \(\WS (n+1)\) and \(\WS (n+2)\), was found and proved
by Romain Ageron.

@@

@@

We give here an intuitive explanation of the above theorem; a formal proof can be found in the appendix. 
Let \((p, q) \in \mathbb{N}^2\) such that there exists a partition of \([\![1,q]\!]\) into \(n\) weakly sum-free
subsets and a partition of \([\![1,p]\!]\) into \(k\) sum-free subsets. Let \(a \in \mathbb{N}\) with \(a > q\) 
and let's try to build a colouring of \([\![1, ap + q]\!]\) into \(n + k\) weakly sum-free subsets. Let 
\(l = a - b - 1\), \(r \in [\![1,q]\!]\) and \(w = a - l - r - 1 = b - r\).

@@

\underline{\textbf{\(\mathcal{C}\) block}} \\
In this block,  we use colors \(1, ..., n\). We colour an integer \(x\) according to its column number (written \(\tilde{\pi}(x)\), it is linked to the
projection on the first line, written \(\pi\), by the relation \(\tilde{\pi}(x) = \pi(x) - a\)). A simple solution is to colour \(x\) with the same colour
as \(\tilde{\pi}(x)\) in the weakly sum-free colouring of \([\![1,q]\!]\). As long as \(2b \leqslant a + r\) (not going two far to the right) and there
is no \(x \in \tilde{\pi}(\mathcal{C})\) such that \(2x \in \tilde{\pi}(\mathcal{C})\) (so that we do not have a sum in \(\mathcal{C}\) when
taking two numbers in the same column), the colours \(1, ..., n\) are sum-free. \\

In particular, taking \(w = l = \lceil \frac{q}{2} \rceil\) and \(r = \lfloor \frac{q}{2} \rfloor\) works, thus obtaining the above theorem.

\begin{remark}
This formula includes the results of Rowley \cite{RowleyWS} as a special case. For \(n>2\),
this formula does not give new lower bounds but in the same way as we introduced \(S^+\)\hyperref[SE]{(Definition
3.1)}, we define \(\WS^+\) and find inequalities between \(\WS^+\),\(\WS\) and \(S\).
\end{remark}

@@

In this subsection, we define the objects and prove the results needed for the general theorem regarding templates for
weak Schur numbers.

@@

Let \((a,b) \in (\mathbb{N}^*)^2\),\(a>b\), \(n \in \mathbb{N}^*\), we will define \(\pi_{a,b}\) the projection:
\[ \pi_{a,b}:x \longmapsto (Id+a\mathbf{1}_{ [\![0,b]\!]})(x \mod a))\]
\end{definition}

@@

\begin{align*}
 \pi(x+y) & = (Id+a\mathbf{1}_{ [\![0,b]\!]})(x+y \mod a)\\
& = (Id+a\mathbf{1}_{ [\![0,b]\!]})(x+\pi(y) \mod a)\text{ since } \pi(y)\=y \mod a \\
& =x+ \pi(y)
\end{align*}

if \(x+\pi(y)\geqslant a\):
\begin{align*}
 \pi(x+y) & = (Id+a\mathbf{1}_{ [\![0,b]\!]})(x+y \mod a)\\
& = (Id+a\mathbf{1}_{ [\![0,b]\!]})(x+\pi(y)\mod a)\text{ since } \pi(y)=y \mod a \\
& = (Id+a\mathbf{1}_{ [\![0,b]\!]})(x+\pi(y) - a) \\
& = x+\pi(y) - a +a\mathbf{1}_{ [\![0,b]\!]})(x+\pi(y) - a) \\
& = x+\pi(y) - a +a \text{ \quad since } x+\pi(y) - a \in [\![0,b]\!] \\
& = x+\pi(y)
\end{align*}

@@

\begin{align*}
 \\\pi(\pi(x)+\pi(y)) & = (Id+a\mathbf{1}_{ [\![0,b]\!]})(\pi(x)+\pi(y)\text{ mod a})\\
& = (Id+a\mathbf{1}_{ [\![0,b]\!]})((Id+a\mathbf{1}_{ [\![0,b]\!]})(x \mod a)+(Id+a\mathbf{1}_{ [\![0,b]\!]})(y \mod a)\ mod a) \\
& = (Id+a\mathbf{1}_{ [\![0,b]\!]})(((x \mod a) + (y \mod a)) \mod a)\\
& =(Id+a\mathbf{1}_{ [\![0,b]\!]})( x+y \mod a)\\
& =\pi(x+y)
\end{align*}

@@

\begin{proposition}
Let \((a,b)\in (\mathbb{N}^*)^2\), \(a>b\), let  \(x\in \mathbb{N}^*\), \(x=a\lambda(x)+\pi(x)-a\)
\end{proposition}
\textsc{Proof :}Let \((a,b)\in (\mathbb{N}^*)^2\), \(a>b\), let  \(x\in \mathbb{N}^*\),
\\\\ \(a\lambda(x)+\pi(x)-a=a\left\lfloor\dfrac{x-b-1}{a}\right\rfloor+(\text{x mod a})+\mathbf{1}_{ [\![0,b]\!]}(x \mod a)\)

@@

if \(x \mod a \leqslant b\):
\(a\lambda(x)+\pi(x)-a=a(\left\lfloor\dfrac{x}{a}\right\rfloor-1)+\text{x mod a}+a=x\)

@@

\subsection{Lower bound for Weak schur numbers using Schur and Weak Schur template numbers}

\begin{theorem}
Let \((q,n,b) \in (\mathbb{N}^*)^3\), let \( (p,k) \in (\mathbb{N}^*)^2\). If there exists a partition of \(k\)
sum-free subsets of \([\![1,p]\!]\) and a partition of \(n\) subsets \((A_1,....,A_n)\) of \([\![1, q]\!]\) which is a
\(b\)-WSF of \(n\) colors and lenght \(q\),
 then there exists a partition of \([\![1, b+p \times (q-b)]\!]\) into \((k+n-1)\) weakly sum-free subsets.
\end{theorem}

@@

\begin{proposition}
Let \((b, k, p) \in \mathbb{N}^*)^3\) and let \(f\) be a colouring associated to a \(b\)-WSF-template of length \(p\) with \(k\) colors. Let
\(c \in \mathbb{N}\) (\(c = \min (A_{k+1} \backslash [\![1, b ]\!]) - 1\) works) and assume there there exists a colouring \(g\) of
\([\![b + 1, b + c]\!]\) with \(k\) colors such that:
\begin{itemize}
	\item \(\forall c \in [\![1, k]\!], \forall (x, y) \in  [\![1, a + b]\!] \times  [\![b + 1, a + b]\!], (f(x) = f(y) \text{ and } \pi(x + y) \leqslant b + c)
	\implies g(\pi(x + y)) \neq f(x)\)
	\item \(\forall c \in [\![1, k]\!], \forall (x, y) \in  [\![1, a + b]\!] \times  [\![b + 1, b + c]\!],  (f(x) = g(y) \text{ and } \pi(x + y) \leqslant b + c)
\implies g(\pi(x + y)) \neq f(x)\)
\end{itemize}
Then, for every \(n \in \mathbb{N}^*\), by using on the last row the colouring \(i \longmapsto g(i - p S(n)\), we have\\
\[ \WS(n+k) \geqslant \WS^+(k+1) S(n) + b + c\]
\end{proposition}

@@

Define \( f : [\![1,b+pa]\!] \longrightarrow [\![1,n]\!] \) such that:
\\\\-if  \(x\leqslant b \text{ (we will note }x \in \mathcal{T}): f(x)=g(x)\)
\\\\-if \( x \in [\![1,b+pa]\!] \text{ and } \pi(x) \notin A_n \text{ (we will note x} \in \mathcal{C}): f(x)=g(\pi(x))\)
\\\\-if \( x \in [\![1,b+pa]\!]  \text{ and }\pi(x) \in A_n \text{ (we will note x} \in \mathcal{R}): f(x)=n-1+h(\lambda(x))\)
\\\\
\(f\) is well defined because \(\pi\) is defined for \(x>b\) and \(\forall x \in \![1,b+pa]\!], f(x)\leqslant n+k-1\) because \(h(\lambda(x))\leqslant k\)

We have parted the integers of \(\![1,b+pa]\!]\) in three disjoints subsets \(\mathcal{T},\mathcal{C} \text{ and } \mathcal{R}\).

We have to verify that f induces weakly-sum-free templates:\\
\\
\underline{if (x,y) \(\in (\mathcal{T})^2\) such that \(f(x)=f(y)\), \(x \neq y\), then \(f(x+y)\neq f(x)\)  :}\\
\\\(x+y<a+b\) since \(b<a\) and \(g(x)=f(x)=f(y)=g(y)\).
\\Hence \(f(x+y)=g(x+y)\neq g(x)=f(x)\)
\\\\
\underline{if \((x,y) \in \mathcal{T} \times \mathcal{C}\) such that \(f(x)=f(y)\), \(x \neq y\), then \(f(x+y)\neq f(x)\)  :}\\
We distinguish two cases:

@@

\begin{itemize}
\item If \(\lambda(y)=\lambda(x+y)\),
\\\(g(x)=g(\pi(y))=n \text{ since } g(y)=g(\pi(y))\)
\\Therefore \(g(\pi(x+y))=g(x+\pi(y)) \neq g(x)=n\)(qv previous proposition)
\\Hence \(f(x+y)=g(\pi(x+y))\neq n\)
\item If  \(\lambda(y)\neq \lambda(x+y)\), \(\lambda(y)+1= \lambda(x+y)\)\\
\(n=f(y)=n-1+h(\lambda(y))\). Hence \(h(\lambda(y))=1\).
\\Moreover \(h(1)=1\), therefore \(h(\lambda(y)+1) \neq 1\)
\\if \(\pi(x+y) \in A_n\), \(f(x+y)=n-1+h(\lambda(x+y))>n\)
\end{itemize}

\underline{if \((x,y) \in (\mathcal{C})^2\) such that \(f(x)=f(y)\), \(x \neq y\), then \(f(x+y)\neq f(x)\)  :}\\
\\Then  \(g(\pi(x))=f(x)=f(y)=g(\pi(y))\). We distinguish two cases:

@@

\begin{align*}
 \\\text{if f(x+y)\(\geqslant\)n, }f(x+y) & =n-1+h(\lambda(x+y))\\
& =n-1+h(\lambda(x)+\lambda(y)) \\
& \neq n-1+h(\lambda(x))\\
& =f(x)
\end{align*}

@@

\subsection{New lower bounds for Weak Schur numbers}
Having found suitable templates, which can be found in the appendix, with a SAT solver, we claim that for all \(n \in
\mathbb{N}^*\):

@@

In this section, we first gave a new construction which can be seen as an equivalent for weakly sum-free partitions of Abott 
and Hanson's construction for sum-free partitions. We then  introduced WSF-templates and generalized this construction. This 
allows us to find new lower bounds and new inequalities for weak Schur numbers. One may notice the significant difference 
between the former lower bounds for weak Schur numbers obtained by conducting a computer search and the new lower bounds 
obtained with WSF-templates (including Rowley's two inequalities). In the next section, we try to analyze this phenomenon.

>>>>

\qquad In this section, we generalize Rowley's constructions for weak Schur numbers \cite{RowleyWS} and give an analogous
for weak Schur numbers of Abbott and Hanson's construction for Schur numbers. By analogy with the previous section,
we then introduce WSF-templates as well as the sequence \(\WS^+(n)\). We find suitable templates and use them to
provide new lower bounds for weak Schur numbers.

@@

\qquad Up to now, there was no equivalent for weak Schur numbers of Abbott and Hanson's construction for Schur numbers
\cite{AbbottHanson}. Here we give a general lower bound for weak Schur numbers as a function of both regular and
weak Schur numbers. The following theorem, inspired by Rowley's inequalities for \(\WS (n+1)\) and \(\WS (n+2)\),
was found and proved by Romain Ageron.

@@

\begin{remark}
This formula includes the results of Rowley \cite{RowleyWS} as a special case. For \(n>2\), this formula does not give
new lower bounds.
\end{remark}

@@

Given that this theorem will appear as a particular case of a more general theorem after the introduction of
templates for weak Schur numbers, we only give here an intuitive explanation of the above theorem; a formal
proof can be found in the appendix.

Let \((p, q) \in \mathbb{N}^2\) such that there exists a partition of \([\![1,q]\!]\) into \(n\) weakly sum-free
subsets and a partition of \([\![1,p]\!]\) into \(k\) sum-free subsets. Let \(a \in \mathbb{N}\) with \(a > q\)
and let's try to build a colouring of \([\![1, ap + q]\!]\) into \(n + k\) weakly sum-free subsets. Let
\(l = a - b - 1\), \(r \in [\![1,q]\!]\) and \(w = a - l - r - 1 = b - r\).

@@

\underline{\textbf{\(\mathcal{C}\) block}} \\
In this block,  we use colors \(1, ..., n\). We colour an integer \(x\) according to its column number (written \(\tilde{\pi}(x)\), it is linked to the
projection on the first line, written \(\pi\), by the relation \(\tilde{\pi}(x) = \pi(x) - a\)). A simple solution is to colour \(x\) with the same colour
as \(\tilde{\pi}(x)\) in the weakly sum-free colouring of \([\![1,q]\!]\). As long as \(2b \leqslant a + r\) (not going two far to the right) and there
is no \(x \in \tilde{\pi}(\mathcal{C})\) such that \(2x \in \tilde{\pi}(\mathcal{C})\) (so that we do not have a sum in \(\mathcal{C}\) when
taking two numbers in the same column), the colours \(1, ..., n\) are sum-free.

In particular, taking \(w = l = \lceil \frac{q}{2} \rceil\) and \(r = \lfloor \frac{q}{2} \rfloor\) works, thus obtaining the above theorem.\\
\par
As in the previous section, we now introduce WSF-templates and the sequence \(\WS^+\) in order to generalize the above construction.

@@

\qquad In this subsection, we introduce WSF-templates and define the objects and prove the results needed
for the general theorem regarding templates for weak Schur numbers.

@@

Let \((a,b) \in (\mathbb{N}^*)^2\),\(a>b\), \(n \in \mathbb{N}^*\), we will define \(\pi_{a,b}\) the projection:
\[ \pi_{a,b}:x \longmapsto (Id+a\mathds{1}_{ [\![0,b]\!]})(x \mod a))\]
\end{definition}

@@

\begin{align*}
 \pi(x+y) & = (Id+a\mathds{1}_{ [\![0,b]\!]})(x+y \mod a)\\
& = (Id+a\mathds{1}_{ [\![0,b]\!]})(x+\pi(y) \mod a)\text{ since } \pi(y)=y \mod a \\
& =x+ \pi(y)
\end{align*}

if \(x+\pi(y)\geqslant a\):
\begin{align*}
 \pi(x+y) & = (Id+a\mathds{1}_{ [\![0,b]\!]})(x+y \mod a)\\
& = (Id+a\mathds{1}_{ [\![0,b]\!]})(x+\pi(y)\mod a)\text{ since } \pi(y)=y \mod a \\
& = (Id+a\mathds{1}_{ [\![0,b]\!]})(x+\pi(y) - a) \\
& = x+\pi(y) - a +a\mathds{1}_{ [\![0,b]\!]})(x+\pi(y) - a) \\
& = x+\pi(y) - a +a \text{ \quad since } x+\pi(y) - a \in [\![0,b]\!] \\
& = x+\pi(y)
\end{align*}

@@

\begin{align*}
 \\\pi(\pi(x)+\pi(y)) & = (Id+a\mathds{1}_{ [\![0,b]\!]})(\pi(x)+\pi(y)\text{ mod a})\\
& = (Id+a\mathds{1}_{ [\![0,b]\!]})((Id+a\mathds{1}_{ [\![0,b]\!]})(x \mod a)+(Id+a\mathds{1}_{ [\![0,b]\!]})(y \mod a)\ mod a) \\
& = (Id+a\mathds{1}_{ [\![0,b]\!]})(((x \mod a) + (y \mod a)) \mod a)\\
& =(Id+a\mathds{1}_{ [\![0,b]\!]})( x+y \mod a)\\
& =\pi(x+y)
\end{align*}

@@

\begin{proposition}
Let \((a,b)\in (\mathbb{N}^*)^2\), \(a>b\), let  \(x\in \mathbb{N}^*\), \(x=a\lambda(x)+\pi(x)-a\)
\end{proposition}
\textsc{Proof :}Let \((a,b)\in (\mathbb{N}^*)^2\), \(a>b\), let  \(x\in \mathbb{N}^*\),
\\\\ \(a\lambda(x)+\pi(x)-a=a\left\lfloor\dfrac{x-b-1}{a}\right\rfloor+(x \mod a)+\mathds{1}_{ [\![0,b]\!]}(x \mod a)\)

@@

if \(x \mod a \leqslant b\):
\(a\lambda(x)+\pi(x)-a=a \left( \left \lfloor \dfrac{x}{a} \right \rfloor - 1 \right)+\text{x mod a}+a=x\)

@@

We now proceed to state and prove the main result of this article.

\subsection{Construction of weak Schur partitions using WSF-templates}

\begin{theorem}
Let \((q,n,b) \in (\mathbb{N}^*)^3\), let \( (p,k) \in (\mathbb{N}^*)^2\). If there exists a partition of \(k\)
sum-free subsets of \([\![1,p]\!]\) and a partition of \(n\) subsets \((A_1,....,A_n)\) of \([\![1, q]\!]\) which is a
\(b\)-WSF-template of \(n\) colors and lenght \(q\),
 then there exists a partition of \([\![1, b+p \times (q-b)]\!]\) into \((k+n-1)\) weakly sum-free subsets.
\end{theorem}

@@

\begin{proposition}
Let \((b, k, p) \in \mathbb{N}^*)^3\) and let \(f\) be a colouring associated to a \(b\)-WSF-template of length \(p\) with \(k\) colors. Let
\(c \in \mathbb{N}\) (\(c = \min (A_{k+1} \backslash [\![1, b ]\!]) - 1\) works) and assume there there exists a colouring \(g\) of
\([\![b + 1, b + c]\!]\) with \(k\) colors such that:

\begin{itemize}
	\item \(\forall c \in [\![1, k]\!], \forall (x, y) \in  [\![1, a + b]\!] \times  [\![b + 1, a + b]\!], (f(x) = f(y) \text{ and } \pi(x + y) \leqslant b + c)
	\implies g(\pi(x + y)) \neq f(x)\)
	\item \(\forall c \in [\![1, k]\!], \forall (x, y) \in  [\![1, a + b]\!] \times  [\![b + 1, b + c]\!],  (f(x) = g(y) \text{ and } \pi(x + y) \leqslant b + c)
\implies g(\pi(x + y)) \neq f(x)\)
\end{itemize}

Then, for every \(n \in \mathbb{N}^*\), by using on the last row the colouring \(i \longmapsto g(i - p S(n)\), we have\\
\[ \WS(n+k) \geqslant \WS^+(k+1) S(n) + b + c\]
\end{proposition}

@@

Define \( f : [\![1,b+pa]\!] \longrightarrow [\![1,n]\!] \) as follows:

\begin{itemize}
\item if  \(x\leqslant b \text{ (we will note }x \in \mathcal{T}): f(x)=g(x)\)
\item if \( x \in [\![1,b+pa]\!] \text{ and } \pi(x) \notin A_n \text{ (we will note x} \in \mathcal{C}): f(x)=g(\pi(x))\)
\item if \( x \in [\![1,b+pa]\!]  \text{ and }\pi(x) \in A_n \text{ (we will note x} \in \mathcal{R}): f(x)=n-1+h(\lambda(x))\)
\end{itemize}

\(f\) is well defined since \(\pi\) is defined for \(x>b\) and \(\forall x \in \![1,b+pa]\!], f(x)\leqslant n+k-1\) because \(h(\lambda(x))\leqslant k\)

We have parted the integers of \([\![1,b+pa]\!]\) in three disjoints subsets \(\mathcal{T},\mathcal{C} \text{ and } \mathcal{R}\).

We have to verify that f induces weakly-sum-free templates:\\
\\
\underline{if \((x,y) \in \mathcal{T}^2\) such that \(f(x)=f(y)\), \(x \neq y\), then \(f(x+y)\neq f(x)\)  :}\\
\\\(x+y<a+b\) since \(b<a\) and \(g(x)=f(x)=f(y)=g(y)\).
\\Hence \(f(x+y)=g(x+y)\neq g(x)=f(x)\)
\\\\
\underline{if \((x,y) \in \mathcal{T} \times \mathcal{C}\) such that \(f(x)=f(y)\), \(x \neq y\), then \(f(x+y)\neq f(x)\)  :}\\
We distinguish two cases:

@@

\begin{itemize}
\item If \(\lambda(y)=\lambda(x+y)\),
\\\(g(x)=g(\pi(y))=n \text{ since } g(y)=g(\pi(y))\)
\\Therefore \(g(\pi(x+y))=g(x+\pi(y)) \neq g(x)=n\) (qv previous proposition)
\\Hence \(f(x+y)=g(\pi(x+y))\neq n\)
\item If  \(\lambda(y)\neq \lambda(x+y)\), \(\lambda(y)+1= \lambda(x+y)\)\\
\(n=f(y)=n-1+h(\lambda(y))\). Hence \(h(\lambda(y))=1\).
\\Moreover \(h(1)=1\), therefore \(h(\lambda(y)+1) \neq 1\)
\\if \(\pi(x+y) \in A_n\), \(f(x+y)=n-1+h(\lambda(x+y))>n\)
\end{itemize}

\underline{if \((x,y) \in \mathcal{C}^2\) such that \(f(x)=f(y)\), \(x \neq y\), then \(f(x+y)\neq f(x)\)  :}\\
\\Then  \(g(\pi(x))=f(x)=f(y)=g(\pi(y))\). We distinguish two cases:

@@

\begin{align*}
 \\\text{if \(f(x+y) \geqslant n\), }f(x+y) & =n-1+h(\lambda(x+y))\\
& =n-1+h(\lambda(x)+\lambda(y)) \\
& \neq n-1+h(\lambda(x))\\
& =f(x)
\end{align*}

@@

\subsection{New lower bounds for Weak Schur numbers}

\qquad Having found suitable templates, which can be found in the appendix, with a SAT solver, we claim that for all \(n \in
\mathbb{N}^*\):

@@

\qquad In this section, we first gave a new construction which can be seen as an equivalent for weakly sum-free partitions of Abott
and Hanson's construction for sum-free partitions. We then  introduced WSF-templates and generalized this construction. This
allows us to find new lower bounds and new inequalities for weak Schur numbers. One may notice the significant difference
between the former lower bounds for weak Schur numbers obtained by conducting a computer search and the new lower bounds
obtained with WSF-templates (including Rowley's two inequalities). In the next section, we try to analyze this phenomenon.
